A Typed Programmatic Interface to Contracts on the Blockchain

### 2021-02-17 papers/execution-model/model.tex
snapshot (with subtyping)

@@

Grammar of types

Typing rules, relying on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language and
$\JTypeValue\STRING\TYPE$ for serialized values as stored on the blockchain.
\begin{mathpar}
  \inferrule{}{\JTypeExpr\TEnv\INT\TINT}

@@

  \inferrule{}{\JTypeExpr\TEnv\PUH{\TADDR}}

@@

  \inferrule{}{\JTypeExpr\TEnv{\STATUSINCLUDING (\INT)}\TSTATUS}

@@

@@

  \inferrule{
    \JTypeExpr\TEnv\EXPR\TYPE \\
    \JTypeExpr\TEnv{\EXPR'}{\TYPE'}
  }{
    \JTypeExpr\TEnv{(\EXPR, \EXPR')}{\TPAIR\ \TYPE\ \TYPE'}
  }

  \inferrule{
    \JTypeExpr\TEnv{\EXPR}{\TPAIR\ \TYPE\ \TYPE'}
  }{
    \JTypeExpr\TEnv{\FST\ \EXPR}\TYPE
  }

  \inferrule{
    \JTypeExpr\TEnv{\EXPR}{\TPAIR\ \TYPE\ \TYPE'}
  }{
    \JTypeExpr\TEnv{\SND\ \EXPR}{\TYPE'}
  }
  \\\\
  \inferrule{
    \JTypeExpr\TEnv\EXPR\TEXCEPTION
  }{
    \JTypeExpr\TEnv{\RAISE\ \EXPR}\TYPE
  }

  \inferrule{
    \JTypeExpr\TEnv\EXPR\TYPE \\
    \JTypeExpr\TEnv{\EXPR'}{\TEXCEPTION\to\TYPE}
  }{
    \JTypeExpr\TEnv{\TRY\ \EXPR\ \EXCEPT\ \EXPR'}\TYPE
  }
  \\\\
  % too generic
  % \inferrule{
  %   \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
  %   \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
  %   \JTypeExpr\TEnv{\EXPR_3}\TADDR \\
  %   \JTypeExpr\TEnv{\EXPR_4}\TSTRING \\
  %   \JTypeExpr\TEnv{\EXPR_5}\TTEZ
  % }{
  %   \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
  % }
  %   
  \inferrule{
    \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
    \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
    % \JTypeExpr\TEnv{\EXPR_3}\TPUK \\
    \JTypeExpr\TEnv{\EXPR_3}{\TCONTRACT\ \TUNIT} \\
    % \JTypeExpr\TEnv{\EXPR_4}\TSTRING \\
    \JTypeExpr\TEnv{\EXPR_5}\TTEZ
  }{
    \JTypeExpr\TEnv{\TRANSFER{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
  }

@@

  \inferrule{
    \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
    \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
    \JTypeExpr\TEnv{\EXPR_3}{\TCODE\ \TYPE_p\ \TYPE_s} \\
    \JTypeExpr\TEnv{\EXPR_4}\TYPE_s \\ % really \TSTRING \\
    \JTypeExpr\TEnv{\EXPR_5}\TTEZ
  }{
    \JTypeExpr\TEnv{\ORIGINATE{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_4}{\EXPR_5}}\TOPH 
  }
\end{mathpar}

Types for queries
\begin{mathpar}
  \inferrule{
    \JTypeExpr\TEnv\EXPR\TPUK
  }{
    \JTypeExpr\TEnv {\QRY\ \GETCOUNTER\EXPR}\TINT
  }

@@

  \inferrule{
    \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  }{
    \JTypeExpr\TEnv {\QRY\ \GETSTORAGE\EXPR}\TSTRING
  }

  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  % }{
  %   \JTypeExpr\TEnv {\QRY\ \GETCODE\EXPR}?
  % }
  %   
  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  % }{
  %   \JTypeExpr\TEnv {\QRY\ \GETTYPE\EXPR}?
  % }
  %   
  \inferrule{
    \JTypeExpr\TEnv\EXPR{\TOPH}
  }{
    \JTypeExpr\TEnv {\QRY\ \GETCONTRACT\EXPR}\TPUH
  }
\end{mathpar}

>>>>

{[Ultra-brief intro to blockchain]}

@@

Subtyping
\begin{mathpar}
  \inferrule{}{\TPUH \SubType \TADDR}

  \inferrule{}{\TPUK \SubType \TADDR}

  \inferrule{}{\TCONTRACT\ \TYPE \SubType \TPUH}

  \inferrule{\TYPEU' \SubType \TYPE' \\ \TYPE \SubType \TYPEU}{
    \TYPE' \to \TYPE \SubType \TYPEU' \to \TYPEU}
\end{mathpar}

Figure~\ref{fig:typing-expressions} presents selected typing rules for
expressions. We rely on an external typing judgment
$\JTypeCode\CODE\TYPE$ for the contract language, which we leave
unspecified, and $\JTypeValue\STRING\TYPE$ for serialized values as
stored on the blockchain. The latter judgment states $\STRING$ is
a string parseable as a value of type $\TYPE$.
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{}{\JTypeExpr\TEnv\INT\TINT}

@@

    \inferrule{}{\JTypeExpr\TEnv\PUH{\TPUH}}

@@

    \inferrule{\JTypeExpr\TEnv\EXPR\TINT}{\JTypeExpr\TEnv{\STATUSINCLUDING
        (\EXPR)}\TSTATUS}

@@

    \inferrule{\JTypeExpr\TEnv\EXPR\TYPE \\ \TYPE \SubType
      \TYPEU}{\JTypeExpr\TEnv\EXPR\TYPEU}

@@

    \inferrule{
      \JTypeExpr\TEnv\EXPR\TYPE \\
      \JTypeExpr\TEnv{\EXPR'}{\TYPE'} }{ \JTypeExpr\TEnv{(\EXPR,
        \EXPR')}{\TPAIR\ \TYPE\ \TYPE'} }
  % 
    % \inferrule{
    % \JTypeExpr\TEnv{\EXPR}{\TPAIR\ \TYPE\ \TYPE'}
    % }{
    %   \JTypeExpr\TEnv{\FST\ \EXPR}\TYPE
    % }
  %   
    %   \inferrule{
    %   \JTypeExpr\TEnv{\EXPR}{\TPAIR\ \TYPE\ \TYPE'}
    % }{
    %   \JTypeExpr\TEnv{\SND\ \EXPR}{\TYPE'}
    % }

    \inferrule{ \JTypeExpr\TEnv\EXPR\TEXCEPTION }{
      \JTypeExpr\TEnv{\RAISE\ \EXPR}\TYPE }

    \inferrule{
      \JTypeExpr\TEnv\EXPR\TYPE \\
      \JTypeExpr\TEnv{\EXPR'}{\TEXCEPTION\to\TYPE} }{
      \JTypeExpr\TEnv{\TRY\ \EXPR\ \EXCEPT\ \EXPR'}\TYPE }
  \end{mathpar}
  \caption{Typing rules for expressions}
  \label{fig:typing-expressions}
\end{figure}
\begin{figure}[tp]
  \begin{mathpar}
    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}\TPUK \\
      % \JTypeExpr\TEnv{\EXPR_3}{\TCONTRACT\ \TUNIT} \\
      \JTypeExpr\TEnv{\EXPR_4}\TUNIT \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
    }

@@

    \inferrule{
      \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
      \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
      \JTypeExpr\TEnv{\EXPR_3}{\TCODE\ \TYPE_p\ \TYPE_s} \\
      \JTypeExpr\TEnv{\EXPR_4}\TYPE_s \\ % really \TSTRING \\
      \JTypeExpr\TEnv{\EXPR_5}\TTEZ }{
      \JTypeExpr\TEnv{\ORIGINATE{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_4}{\EXPR_5}}\TOPH
    }
  \end{mathpar}
\begin{mathpar}
  \inferrule{
    \JTypeExpr\TEnv\EXPR\TPUK
  }{
    \JTypeExpr\TEnv {\QRY\ \GETCOUNTER\EXPR}\TINT
  }

@@

  \inferrule{
    \JTypeExpr\TEnv\EXPR{\TPUH}
  }{
    \JTypeExpr\TEnv {\QRY\ \GETSTORAGE\EXPR}\TSTRING
  }

  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  % }{
  %   \JTypeExpr\TEnv {\QRY\ \GETCODE\EXPR}?
  % }
  %   
  % \inferrule{
  %   \JTypeExpr\TEnv\EXPR{\CONTRACT\ \TYPE}
  % }{
  %   \JTypeExpr\TEnv {\QRY\ \GETTYPE\EXPR}?
  % }
  %   
  \inferrule{
    \JTypeExpr\TEnv\EXPR{\TOPH}
  }{
    \JTypeExpr\TEnv {\QRY\ \GETCONTRACT\EXPR}\TPUH
  }
\end{mathpar}
  \caption{Typing rules for blockchain operations and queries}
  \label{fig:typing-blockchain-operations}
\end{figure}
% Generic rule for transfers
% \begin{mathpar}
%  \inferrule{
%    \JTypeExpr\TEnv{\EXPR_1}\TTEZ \\
%    \JTypeExpr\TEnv{\EXPR_2}\TPUK \\
%    \JTypeExpr\TEnv{\EXPR_3}\TADDR \\
%    \JTypeExpr\TEnv{\EXPR_4}\TSTRING \\
%    \JTypeExpr\TEnv{\EXPR_5}\TTEZ
%  }{
%    \JTypeExpr\TEnv{\TRANSFER[\EXPR_4]{\EXPR_1}{\EXPR_2}{\EXPR_3}{\EXPR_5}}\TOPH
%  }
% \end{mathpar}
